Case Study: RNA-Seq Differential Analysis (mir200c knockdown)

### 2018-03-22 datasets/GWAS/GWAS.Rmd
Fix a couple typos in RNAseq Rmds

```{r, wkspace-setup, results='hide', message=FALSE, warning=FALSE}
# data and results directories
system("mkdir -p DATA")
system("mkdir -p RESULTS")
datdir <- "DATA/"
resdir <- "RESULTS/"

# wrangling & plotting tools
library(data.table)
library(readxl)
library(readr)
library(dplyr)
library(ggplot2)
library(cowplot)

# comparison tools/functions
library(R.utils)
sourceDirectory("../R/")

# set up parallel backend
library(BiocParallel)
cores <- 8
multicoreParam <- MulticoreParam(workers = cores)
```

@@

```{r, GWAS1-download}
if(!file.exists(paste0(datdir,"BMI.SNPadjSMK.CombinedSexes.EuropeanOnly.txt"))){
  download.file(url="http://portals.broadinstitute.org/collaboration/giant/images/3/3a/BMI.SNPadjSMK.zip", 
             destfile=paste0(datdir,"BMI.SNPadjSMK.zip")) 
  unzip(paste0(datdir,"BMI.SNPadjSMK.zip"), exdir=datdir)
  file.remove(paste0(datdir,"BMI.SNPadjSMK.zip"))

  dfiles <- list.files(path=datdir, pattern="BMI.SNPadjSMK.*.txt", 
                     full.names = TRUE)
  dfiles <- dfiles[!grepl("BMI.SNPadjSMK.CombinedSexes.EuropeanOnly.txt", dfiles)]
  file.remove(dfiles)
}
```

@@

```{r, GWAS1-refdata}
reffile <- paste0(datdir, "1000G_20101123_v3_GIANT_chr1_23_minimacnamesifnotRS_CEU_MAF0.01")
if (!file.exists(paste0(reffile, ".fam"))){
  download.file(url="http://neurogenetics.qimrberghofer.edu.au/iSECA/1000G_20101123_v3_GIANT_chr1_23_minimacnamesifnotRS_CEU_MAF0.01.zip", 
              destfile=paste0(reffile, ".zip"))
  unzip(paste0(reffile, ".zip"))
  file.remove(paste0(reffile, ".zip"))
}
```

@@

```{r, GWAS1-verify-contents}
bmi <- fread(paste0(datdir, "BMI.SNPadjSMK.CombinedSexes.EuropeanOnly.txt"),
             header=TRUE)
dim(bmi)
head(bmi)
```

@@

```{r, GWAS1-snplist, eval=TRUE}
# load reference data 
onekg <- fread(paste0(datdir, "1000G_20101123_v3_GIANT_chr1_23_minimacnamesifnotRS_CEU_MAF0.01.bim"))$V2

# construct input list of SNPs that overlap - 97.5% overlap
bmi$BP <- as.numeric(unlist(lapply(strsplit(bmi$markername, ":"), function(x) x[[2]])))
write_delim(bmi %>% dplyr::rename(CHR=chromosome, SNP=rs_id, A1=allele_1, A2=allele_2) %>%
              mutate(P=0.5, F_A=Freq_Allele1_HapMapCEU, F_U=F_A, CHISQ=1, OR=1) %>% 
              select(CHR, SNP, BP, A1, F_A, F_U, A2, CHISQ, P, OR) %>%
              filter(SNP %in% onekg), 
            path=paste0(datdir, "rslist.assoc"), delim="\t")

@@

```{bash, clump, eval=FALSE}
# use PLINK clumping tool to obtain a set of independent SNPs by LD
plink \
--bfile DATA/1000G_20101123_v3_GIANT_chr1_23_minimacnamesifnotRS_CEU_MAF0.01 \
--clump DATA/rslist.assoc \
--clump-field P \
--clump-p1 0.9999 \
--clump-p2 0.9999 \
--clump-r2 0.2 \
--clump-kb 250 \
--out DATA/plink_clump
```

@@

```{r, GWAS1-prune}
clump <- fread(paste0(datdir, "plink_clump.clumped"))

# subset to include only the SNPs in the ld.list
bmi <- bmi[bmi$rs_id %in% clump$SNP,]

@@

duration <- NA
if (!file.exists(resN)){
  t1 <- proc.time()
  sbN <- bd %>% buildBench(data=(bmi %>% mutate(ind_covariate=ind_covar_N)), 
                          ftCols = c("ind_covar_N"),
                          parallel=TRUE, BPPARAM=multicoreParam)
  metadata(sbN)$data_download_link <- 
    "http://portals.broadinstitute.org/collaboration/giant/images/3/3a/BMI.SNPadjSMK.zip"
  saveRDS(sbN, file=resN)
  duration <- round((proc.time()-t1)[3]/60,1)
}else{
  sbN <- readRDS(resN)
}
```

@@

# properties of method-exclusive discoveries
methods <- c("ashs", "lfdr", "ihw-a10", "bl-df03", "qvalue", "bh", "bonf",
             "scott-empirical", "scott-theoretical")
mcols(sbN)$ind_covariate <- mcols(sbN)$ind_covar_N
plotCovariateBoxplots(sbN, alpha=0.1, nsets=6, methods=methods, trans=log2, maxNum=3) +
  ggtitle("Covariate 1: Sample size")

```

>>>>

```{r, wkspace-setup, results='hide', message=FALSE, warning=FALSE}
library(data.table)
library(readxl)
library(readr)
library(dplyr)
library(ggplot2)
library(cowplot)
library(BiocParallel)

## load helper functions
for (f in list.files("../R", "\\.(r|R)$", full.names = TRUE)) {
    source(f)
}

# data and results directories
datdir <- "data"
resdir <- "results"
dir.create(datdir, showWarnings = FALSE)
dir.create(resdir, showWarnings = FALSE)

# set up parallel backend
cores <- 8
multicoreParam <- MulticoreParam(workers = cores)
```

@@

```{r, GWAS1-download}
if (!file.exists(file.path(datdir, "BMI.SNPadjSMK.CombinedSexes.EuropeanOnly.txt"))) {
  download.file(url = "http://portals.broadinstitute.org/collaboration/giant/images/3/3a/BMI.SNPadjSMK.zip", 
             destfile = file.path(datdir, "BMI.SNPadjSMK.zip")) 
  unzip(file.path(datdir, "BMI.SNPadjSMK.zip"), exdir = datdir)
  file.remove(file.path(datdir,"BMI.SNPadjSMK.zip"))

  dfiles <- list.files(path = datdir, pattern = "BMI.SNPadjSMK.*.txt", 
                       full.names = TRUE)
  dfiles <- dfiles[!grepl("BMI.SNPadjSMK.CombinedSexes.EuropeanOnly.txt", dfiles)]
  file.remove(dfiles)
}
```

@@

```{r, GWAS1-refdata}
reffile <- file.path(datdir, "1000G_20101123_v3_GIANT_chr1_23_minimacnamesifnotRS_CEU_MAF0.01")
if (!file.exists(paste0(reffile, ".fam"))) {
    download.file("http://neurogenetics.qimrberghofer.edu.au/iSECA/1000G_20101123_v3_GIANT_chr1_23_minimacnamesifnotRS_CEU_MAF0.01.zip", 
                  destfile = paste0(reffile, ".zip"))
    unzip(paste0(reffile, ".zip"), exdir = datdir)
    file.remove(paste0(reffile, ".zip"))
}
```

@@

```{r, GWAS1-verify-contents}
bmi <- fread(file.path(datdir, "BMI.SNPadjSMK.CombinedSexes.EuropeanOnly.txt"),
             header = TRUE)
dim(bmi)
head(bmi)
```

@@

```{r, GWAS1-snplist, eval=TRUE}
# load reference data 
onekg <- fread(file.path(datdir,  "1000G_20101123_v3_GIANT_chr1_23_minimacnamesifnotRS_CEU_MAF0.01.bim"))$V2

# construct input list of SNPs that overlap - 97.5% overlap
bmi$BP <- as.numeric(unlist(lapply(strsplit(bmi$markername, ":"), function(x) x[[2]])))
write_delim(bmi %>%
            dplyr::rename(CHR=chromosome, SNP=rs_id, A1=allele_1, A2=allele_2) %>%
            mutate(P=0.5, F_A=Freq_Allele1_HapMapCEU, F_U=F_A, CHISQ=1, OR=1) %>% 
            select(CHR, SNP, BP, A1, F_A, F_U, A2, CHISQ, P, OR) %>%
            filter(SNP %in% onekg), 
            path=file.path(datdir,  "rslist.assoc"), delim="\t")

@@

```{r, clump}
## use PLINK clumping tool to obtain a set of independent SNPs by LD
if (!dir.exists(plink_clump)) {
    system(paste("plink",
                 "--bfile DATA/1000G_20101123_v3_GIANT_chr1_23_minimacnamesifnotRS_CEU_MAF0.01",
                 "--clump DATA/rslist.assoc",
                 "--clump-field P",
                 "--clump-p1 0.9999",
                 "--clump-p2 0.9999",
                 "--clump-r2 0.2",
                 "--clump-kb 250",
                 "--out DATA/plink_clump"))
}
```

@@

```{r, GWAS1-prune}
clump <- fread(file.path(datdir, "plink_clump.clumped"))

# subset to include only the SNPs in the ld.list
bmi <- bmi[bmi$rs_id %in% clump$SNP, ]

@@

duration <- NA
if (!file.exists(resN)) {
  t1 <- proc.time()
  sbN <- bd %>% buildBench(data=(bmi %>% mutate(ind_covariate=ind_covar_N)), 
                          ftCols = c("ind_covar_N"),
                          parallel=TRUE, BPPARAM=multicoreParam)
  metadata(sbN)$data_download_link <- 
    "http://portals.broadinstitute.org/collaboration/giant/images/3/3a/BMI.SNPadjSMK.zip"
  saveRDS(sbN, file=resN)
  duration <- round((proc.time()-t1)[3]/60,1)
}else{
  sbN <- readRDS(resN)
}
```

```{r, echo=FALSE}
if (!is.na(duration)) {
  message("This step took ", duration, " minutes for ", nrow(bmi), " SNPs using ",
          cores, " cores.")
}
```

@@

# properties of method-exclusive discoveries
methods <- c("ashs", "lfdr", "ihw-a10", "bl-df03", "qvalue", "bh", "bonf",
             "scott-empirical", "scott-theoretical")
mcols(sbN)$ind_covariate <- mcols(sbN)$ind_covar_N
plotCovariateBoxplots(sbN, alpha=0.1, nsets=6, methods=methods, trans=log2, maxNum=3) +
    ggtitle("Covariate 1: Sample size")
```

### 2018-03-22 datasets/RNA-Seq/RNAseq-gtex.Rmd
Fix a couple typos in RNAseq Rmds

We download a DESeqDataSet object containing gene level counts for a subset of the GTEx data. The data set includes females and two cell types, nucleus accumbens and the putamen.

@@

We use DESeq2 to test each gene for differential gene expression between the two cell types. We set the parameter `independentFiltering=FALSE` to skip the independent filtering step, as this step would be redundant with some of the FDR control methods that use gene expression as an independent covariate to increase power.

@@

These discreteness, however, corresponded to genes with very low expression values. For example, there were `r nrow(dplyr::filter(res, ind_covariate <= 1))` genes with less than an average of 1 read across the 20 samples. The distribution of p-values looked much better when removing this very lowly expressed genes.

@@

ash was the method that rejected a larger number of hypothesis, followed by lfdr and scott-theorethical.

@@

```{r, rnaseq-boxplots}
## properties of method-exclusive discoveries
methods <- c("ashs", "lfdr", "ihw-a10", "bl-df03", "qvalue", "bh", "bonf")
plotCovariateBoxplots(sb, alpha = 0.1, nsets = 6,
                      methods = methods, trans = log2, maxNum = 5)
```

>>>>

We download a DESeqDataSet object containing gene level counts for GTEx samples from the nucleus accumbens and the putamen.

@@

We use DESeq2 to test for differential gene expression between the two cell types. We set the parameter `independentFiltering=FALSE` to skip the independent filtering step, as this step would be redundant with some of the FDR control methods that use gene expression as an independent covariate to increase power.

@@

This discreteness, however, corresponded to genes with very low expression values. For example, there were `r nrow(dplyr::filter(res, ind_covariate <= 1))` genes with less than an average of 1 read across the 20 samples. The distribution of p-values looked much better when removing these very lowly expressed genes.

@@

ash was the method that rejected the largest number of hypotheses, followed by lfdr and scott-theoretical.

@@

```{r, rnaseq-boxplots}
methods <- c("ashs", "lfdr", "ihw-a10", "bl-df03", "qvalue", "bh", "bonf")
plotCovariateBoxplots(sb, alpha = 0.1, nsets = 6,
                      methods = methods, trans = log2, maxNum = 5)
```

### 2018-03-22 datasets/RNA-Seq/RNAseq-mir200c.Rmd
Fix a couple typos in RNAseq Rmds

As a second RNA-seq dataset, we will test for differences in gene expression upon the knockout of the microRNA mir-200c [(Kim et al., 2013)](http://dx.doi.org/10.1038/nsmb.2701).  The raw fastq files can be found under the accession number `SRP030475`. As the number of samples is limited the experiment might be underpowered, as in most RNA-seq analysis. This is an experimental scenario that could benefit from power gaining from using modern FDR control methods.

@@

We next subset for samples containing the control samples and the samples where mir200c was knocked down. Recount2 downloads data as a RangeSummarizedExperiment object, so we convert this into a DESeqDataSet object.

@@

As with the GTEx example, the mean counts is an informative covariate.

@@

Similarly to the GTEx dataset, keeping all the tests results in a strange discreteness. This is removed once we filter very lowly expressed genes. For the first covariate bin, however, there is a strange behaviour in which the distribution seems a bit skewed towards larger p-values.

@@

There are some warnings from both `BH` and `scott-empirical`. However, they do return results. I tried to increase the `nmids` parameter but did not help.

@@

Scott empirical rejects many more hypothesis than the rest of the methods, followed by lfdr and IHW.

>>>>

As a second RNA-seq dataset, we will test for differences in gene expression upon the knockout of the microRNA mir-200c [(Kim et al., 2013)](http://dx.doi.org/10.1038/nsmb.2701).  The raw fastq files can be found under the accession number `SRP030475`. As the number of samples is limited the experiment might be underpowered, as in most RNA-seq analysis. This is an experimental scenario that could benefit from power gained using modern FDR control methods.

@@

We next subset for samples containing the control samples and the samples where mir200c was knocked down. recount2 downloads data as a RangeSummarizedExperiment object, so we convert this into a DESeqDataSet object.

@@

As with the GTEx example, the mean counts is used as the informative covariate.

@@

Similar to the GTEx dataset, keeping all the tests results in a strange discreteness.
This is removed once we filter very lowly expressed genes.
For the first covariate bin, however, there is a strange behaviour in which the distribution seems a bit skewed towards larger p-values.

@@

There are some warnings from both `BH` and `scott-empirical`. However, they do return results.
I tried to increase the `nmids` parameter for Scott's FDR Regression but this does not appear to make a difference.

@@

`scott-empirical` rejects many more hypothesis than the rest of the methods, followed by `lfdr` and `ihw`.
